---
title: "EDS 213 Week 2 Assignment: Data Cleaning"
format: html
author: "Ava Robillard"
date: 4/14/26
---

### Arctic Shorebird Demographics Network (ASDN) `Snow_survey` data cleaning

```{r}
#| message: FALSE
# Load libraries
library(tidyverse)

# Import data
# Snow_cover column pre-cleaned 
snowsurvey <- read_csv(here::here("data", "processed", "snow_cover.csv"))
```

#### Water_cover column

```{r}
# View data types to infer any issues
glimpse(snowsurvey)
```
Based on the glimpse of the file, `Snow_cover` was successfully cleaned and converted to numeric. `Water_cover`, `Land_cover`, and `Total_cover` are still of type character, which is problematic for functioning as percentages as described in the metadata.

To address this, we will explore what is causing these columns to be cast as character.

```{r}
#| warning: FALSE
# Check for anything non-numeric when trying to coerce
snowsurvey %>%
  count(Water_cover) %>% 
  filter(is.na(as.numeric(Water_cover)))
```

The metadata states that in all data fields, a blank or "-" indicates absence of data. Therefore, we can assume that the characters -, ., n/a, and unk all likely refer to unknown values, so can be converted to NA.
 
To double check that we can convert the . values to NA, we can view all of those rows. The other cover columns also contain dots, so they can be converted to NA.

```{r}
# View rows
snowsurvey %>%
  filter(Water_cover == ".") %>%
  View()
```

```{r}
# Fix values by converting to NA
snowsurvey_fixed <- snowsurvey %>% 
  mutate(Water_cover = ifelse(Water_cover %in% c("-", ".", "n/a", "unk"), NA, Water_cover)) %>% 
  # Convert Water_cover column to numeric
  mutate(Water_cover = as.numeric(Water_cover))
```

Now that we have transformed the column to the correct data type, we need to ensure that the expectations for a percentage are followed (0-100)

```{r}
# Check for rows where Water_cover is out of expected bounds
 snowsurvey_fixed %>% 
  filter((Water_cover > 100) | (Water_cover < 0)) 
```
One row has a value of 353, and we can see that the Land_cover column has a large negative value likely to offset this, which we will eventually need to fix as well. We will set this row's value to NA as it is not within the percentage range stated in the metadata.
 
```{r}
# Convert high values to NA
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Water_cover = ifelse(Water_cover > 100, NA, Water_cover))
```


#### Land_cover column

Next, let's check the same key issues in the Land_cover column.

```{r}
#| warning: FALSE
# Check for anything non-numeric when trying to coerce
snowsurvey_fixed %>%
  count(Land_cover) %>% 
  filter(is.na(as.numeric(Land_cover)))
```

We have a very similar distribution of the same character strings preventing this column from being numeric, so we can similary convert them to NAs.

```{r}
# Fix values by converting to NA
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Land_cover %in% c("-", ".", "n/a", "unk"), NA, Land_cover)) %>% 
  # Convert Land_cover column to numeric
  mutate(Land_cover = as.numeric(Land_cover))
```

Now that we have transformed the column to the correct data type, we again need to ensure that the expectations for a percentage are followed (0-100)

```{r}
# Check for rows where Land_cover is out of expected bounds
 snowsurvey_fixed %>% 
  filter((Land_cover > 100) | (Land_cover < 0)) 
```
We have two strange values, -100 and -298. We will set these row's values to NA as they are not within the percentage range stated in the metadata.

We can also see that in these columns, changing these values to NA has caused the three cover types not to add to 100 for the Total_cover column, which we will address next.

```{r}
# Convert low values to NA
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Land_cover < 0, NA, Land_cover))
```

#### Total_cover column

The metadata for this data states that this column serves to check the above percents (snow, water, and land) and should always sum to 100. We will use this to find rows with problematic values and correct them so that the column only returns 100 percents. 

Now that the columns are all numeric, we can add them.

```{r}
# Re-compute Total_cover column after cleaning other rows 
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Total_cover = Snow_cover + Water_cover + Land_cover)
```

```{r}
# check that existing Total_cover column only contains 100s
 snowsurvey_fixed %>% 
  filter(Total_cover != 100)
```

We can see that there are 4,573 rows of data where the recomputed Total_cover column did not add up to 100, so we need to carefully decide how to handle them. 

Consulting the metadata, the Total_cover column states that the total sum should always sum to 100, so we can assume that there is something problematic with these data rows and how the data was collected (not everything is accounted for in snow cover, water cover, and land cover, or there is extra cover above 100 percent which means there was some error in calculating the cover of each). Because we have a total of 42,830 rows in the raw data set, removing these rows would only remove about 10% of the data, which seems reasonable to preserve the integrity of the columns that the metadata expects. 

```{r}
# Filter to only columns that have 100 in the recomputed total
all_cover_fixed_AVAROBILLARD <- snowsurvey_fixed %>%
  filter(Total_cover == 100) 
```

```{r}
# Check unique values in Total_cover 
unique(all_cover_fixed_AVAROBILLARD$Total_cover)
```

```{r}
# Re-check data set types overall 
glimpse(all_cover_fixed_AVAROBILLARD)
```
```{r}
# Re-check data set ranges overall 
summary(all_cover_fixed_AVAROBILLARD)
```

`Snow_cover`, `Water_cover`, `Land_cover` and `Total_cover` are all numeric, and confirmed to be 0-100 and add to 100 for `Total_cover`.

#### Save processed data

```{r}
# write file
write_csv(all_cover_fixed_AVAROBILLARD, here::here("data", "processed", "all_cover_fixed_AVAROBILLARD.csv"))
```

